### 2023-09-03 sections/1-introduction.tex
some intro and lots of discussion work

In particular, $k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Naive implementations of kNN, whose time complexity is linear in the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not 
exploit the geometric and topological structure inherent in large datasets. Approximate search is sufficient for some applications, but the 
need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$.

[TODO: talk about need for exact search in a medical setting; talk to Noah about this, good strains of bacteria, drug response curves]

>>>>

In particular, $k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Naive implementations of kNN, whose time complexity is linear in the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not 
exploit the geometric and topological structure inherent in large datasets. Approximate search is sufficient for some applications, but the 
need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$, and small values of $k$ may be necessary when classes are not well-separated.

For the sake of clarity, it is important to be precise about what we mean by \emph{exact} $k$-NN, and exact search in general.
In the context of $\rho$-NN search, exact search typically means returning only those data points that are identical to the query.
There are many algorithms that perform well here, such as FM-Index~\cite{ferragina2000opportunistic}, which are less applicable to $\rho$-NN search.
In contrast, in the context of $k$-NN search, while all approaches return exactly $k$ elements (ignoring how distance ties are handled), some approaches may miss some of the $k$ nearest elements, and erroneously include some of the elements outside the truly nearest $k$.
In any case, $k$-NN is typically used for inference; the \emph{prediction} of a class for a query is obviously not guaranteed to be correct.
When we refer to \emph{exact} $k$-NN, we refer to precisely the $k$ nearest neighbors being returned; no closer points are elided and no further points are included.
This has implications for the accuracy of classification, but does not guarantee perfect classification accuracy.

There is some evidence that non-metrics such as cosine distance perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance}; this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could be exhibit suboptimal accuracy.

### 2023-09-03 sections/2-methods.tex
some intro and lots of discussion work

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.
In CHESS, each cluster stored a list of indices into the dataset, 
which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices for each of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
Assuming $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost. 
[TODO: talk to Noah about how to phrase this, how to get at that balanced tree is the worst case]

@@

After the first $\log_2{\frac{n}{2k}}$ iterations, we have that the cardinalities of clusters in $Q$ are at most $2k$, at which point, based on our assumption, we have that the length of $Q$ is close to $2d$.

>>>>

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.
In CHESS, each cluster stored a list of indices into the dataset, 
which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices for each of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
Assuming a balanced tree and thus $\mathcal{O}\log_2(n)$ depth, this approach had $\mathcal{O}(n\log_2(n)$ memory cost. We note that while a balanced tree is classically a best-case scenario (e.g. binary search), it represents a worst-case scenario for space complexity in this application; the space complexity of a completely unbalanced (linear) tree would be $\mathcal{O}(n)$.

@@

After the first $\log_2{\frac{n}{2k}}$ iterations, we have that the cardinalities of clusters in $Q$ are at most $2k$, at which point, based on our assumption, we have that the length of $Q$ is close to $2d$.
Therefore, the asymptotic complexity of the sieve algorithms is $\mathcal{O}(log_2(\frac{n}{d})$.

### 2023-09-03 sections/5-discussion.tex
some intro and lots of discussion work

\section{Discussion}
\label{sec:discussion}

Discuss results and future work \dots

CLAM-CAKES.
This is most effective when datasets exhibit low metric entropy and low local fractal dimension.

Unlike LSH, this is extensible to any user provided distance function.

Exactness of search and metric vs non-metric.

While CAKES has perfect recall for distance metrics, it exhibits worse recall under non-metrics (specifically, cosine distance). This bears further investigation, but likely stems from how heavily the knn and clustering algorithms depend on the triangle inequality.

\subsection{Future Works}
\label{subsec:results:future-works}

For future work, we plan to investigate compression by representing differences at each level of the
binary tree, particularly for string or genomic data, where all differences are discrete.
When we descend a level in the cluster tree, it takes fewer bits to encode each point in terms of the center. 
However, each descent of a level in the cluster tree increases the bit-cost of specifying a particular cluster. 
This presents the following tradeoff: \emph{How do we algorithmically determine the depth to which we should cluster a dataset to achieve optimal compression?} 
In our future work, we will explore which properties of the dataset and of clusters can inform this optimal depth.

@@

Additionally, we would like to implement additional distance functions to improve the genericity of CAKES. 
For example, we would like to implement Tanimoto Distance using maximal-common-subgraph for molecular structures. Ultimately, 
we would also like to implement Wasserstein distance for dimensions higher than two. In order to make this possible, 
we plan to implement GPU-acceleration for the more computationally-intensive distance functions like Wasserstein.

@@

>>>>

\section{Discussion and Future Work}
\label{sec:discussion}

We have presented CAKES, a method for fast $k$-nearest-neighbors search that can adapt to a variety of distance functions.
CAKES provides exact $k$-NN when the distance function in use is a metric (particularly, when it obeys the triangle inequality).
CAKES is most effective when datasets are not uniform and do not obey a ``simple'' distribution such as Gaussian, but rather adhere to ``interesting'' low-dimensional manifolds even when embedded in a high-dimensional space; thus, CAKES is best when the manifold hypothesis is upheld.

Unlike locality-sensitive hashing (LSH) approaches, which require a novel hashing function for each distance metric in use, CAKES is extensible to any user-provided distance function.
Also unlike LSH approaches, CAKES guarantees perfect precision and recall as long as the distance function in use is a \emph{metric}; that is, it obeys the properties of identity, symmetry, and the triangle inequality.
As mentioned in Section~\ref{sec:introduction}, this should not be conflated with having $k$-NN itself provide perfect \emph{inference} accuracy.

In comparison to other methods, While CAKES has perfect recall for distance metrics, it exhibits worse recall under non-metrics (specifically, cosine distance). This bears further investigation, but likely stems from how heavily the knn and clustering algorithms depend on the triangle inequality.
Preliminary results suggest that when FAISS and HNSW are tuned for lower recall, they are still slower than CAKES in all cases.

Some of the questions raised by this study suggest that future work is justified; a comparison across more datasets and larger datasets is clearly in order, as is further analysis with distance functions that existing methods such as FAISS do not support, such as Wasserstein distance~\cite{vallender1974calculation} (particularly for arbitrary dimensionality), as Tanimoto distance~\cite{bajusz2015tanimoto}, which is particularly applicable to computing distances based on maximal common subgraphs of molecular structures.
Incorporating these or other distance functions in CAKES only requires providing a Rust implementation of the distance function.

In future work, we plan to investigate compression by representing differences at each level of the
binary tree, particularly for string or genomic data, where all differences are discrete.
When we descend a level in the cluster tree, it takes fewer bits to encode each point in terms of the center. 
However, each descent of a level in the cluster tree increases the bit-cost of specifying a particular cluster. 
This presents the following tradeoff: \emph{How do we algorithmically determine the depth to which we should cluster a dataset to achieve optimal compression?} 
In our future work, we will explore which properties of the dataset and of clusters can inform this optimal depth.

@@

@@

% since this online update takes advantage of CAKES' fast search, it allows us to have our cake and eat it, too.
